---
title: "Classifiers"
author: "Lindsay Hracs"
date: "2024-10-08"
output: html_document
---

## Purpose

This script contains code for a random forest classifier that is used to classify countries at a particular time period as IBD epidemiologic Stage 1, Stage 2, or Stage 3; code is for extension to novel data points.

## Outline

| 1. Script setup
| 2. Install and load packages
| 3. Load data
| 4. Data prep
| 5. Imputation
| 6. Random forest
|       a. Data split
|       b. Training and validation
|       c. Classify unlabelled data
|       d. Output

## 1. Script setup

```{r setup, include = TRUE}

knitr::opts_chunk$set(echo = TRUE)

# Clear environment

rm(list = ls(all = TRUE))
```

## 2. Installl and load packages

```{r required packages}

# Ensure pacman is installed

if (!require("pacman")) install.packages("pacman")

# Install and load packages

pacman::p_load(
      rio,                
      openxlsx,
      tidyverse,          
      ggpubr,
      plotly,
      dplyr,
      htmlwidgets,
      manipulateWidget,
      htmltools,
      RColorBrewer,
      stringr,
      randomForest,
      caret)
         
```

## 3. Load data

```{r load data, warning = FALSE}
# Update file paths as needed

# Incidence and prevalence data
raw <- import("~/file/path/file_name.xlsx", guess_max = 3000) # can ignore warnings for unused cols
 
# Labelled data output from kNN supported iterative data labelling
labelled <- read.csv("~/file/path/file_name_labelled_data.csv")
```

## 4. Data prep

```{r data prep}

# Filter dataset

data <- raw %>%
    filter(age_cat == "All",
           !is.na(CD_rate) | !is.na(UC_IBDu_rate)) %>% 
    group_by(country, name, data_type, year_data) %>%
    mutate(CD_collapsed = mean(CD_rate, na.rm = TRUE),
           UC_collapsed = mean(UC_IBDu_rate, na.rm = TRUE)) %>%
    mutate(IBD_rate = (CD_collapsed + UC_collapsed)) %>%
    distinct(country, name, data_type, year_data, .keep_all = TRUE) %>%
    dplyr::select(country, name, epi_stage2020, year_data, decade, data_type, CD = CD_collapsed, 
                  UC = UC_collapsed, IBD = IBD_rate, study_id) %>%
    ungroup() %>%
    mutate_all(~ifelse(is.nan(.), NA, .)) %>%
    select(country, name, year_data, decade, data_type, CD, UC)

# Format labelled dataset

labelled <- labelled %>%
    rename("stage" = "stage_manual_adj") %>% 
    mutate(status = "labelled") %>%
    select(country, name, year_data, decade, data_type, CD, UC, stage, status)

# Join labelled data with complete dataset

joined <- left_join(data, labelled, by = c("country", "name", "data_type", "year_data"))

# Format joined dataframe

joined <- joined %>%
    rename(c("decade" = "decade.x", "CD" = "CD.x", "UC" = "UC.x", "stage_lab" = "stage", "status_lab" = "status")) %>%
    mutate(stage = ifelse(is.na(stage_lab), "unclassified", stage_lab),
           status = ifelse(is.na(status_lab), "unlabelled", status_lab)) %>%
    select(-decade.y, -CD.y, -UC.y, -stage_lab, -status_lab)

# Create incidence subset with rate of change calculations

inc <- joined %>% 
    filter(data_type == "Incidence") %>%
    group_by(country, name, data_type) %>% # no year in group 
    arrange(year_data, .by_group = TRUE) %>%
    mutate(count = seq_len(n())) %>% # create variable with sequence per group for change calc
    mutate(CD_change = ifelse(count > 1, (CD - lag(CD))/(year_data - lag(year_data)), NA),
           UC_change = ifelse(count > 1, (UC - lag(UC))/(year_data - lag(year_data)), NA)) %>%
    ungroup()

# Create prevalence subset with rate of change calculations

prev <- joined %>% 
    filter(data_type == "Prevalence") %>%
    group_by(country, name, data_type) %>% # no year in group
    arrange(year_data, .by_group = TRUE) %>%
    mutate(count = seq_len(n())) %>% # create variable with sequence per group for change calc
    mutate(CD_change = ifelse(count > 1, (CD - lag(CD))/(year_data - lag(year_data)), NA),
           UC_change = ifelse(count > 1, (UC - lag(UC))/(year_data - lag(year_data)), NA)) %>%
    ungroup()

# Format incidence data frame

inc <- inc %>%
    rename("CD_inc" = "CD", "UC_inc" = "UC", "CD_change_inc" = "CD_change", "UC_change_inc" = "UC_change", "stage_inc" = "stage", "status_inc" = "status") %>%
    select(-data_type, -count)

# Format incidence data frame

prev <- prev %>%
    rename("CD_prev" = "CD", "UC_prev" = "UC", "CD_change_prev" = "CD_change", "UC_change_prev" = "UC_change", "stage_prev" = "stage", "status_prev" = "status") %>%
    select(-data_type, -count)


# Create wide data set because classification is on the basis of inc and prev of CD and UC
    # incidence and prevalence need to be on the same row

wide <- merge(inc, prev, by = c("country", "name", "year_data"), all = TRUE)
```

## 5. Imputation
```{r imputation, warning = FALSE, error = FALSE}

# Create a single stage column and single decade column

combined <- wide %>%
    mutate(stage = coalesce(stage_inc, stage_prev),
           status = coalesce(status_inc, status_prev),
           decade = coalesce(decade.x, decade.y),
           CD_inc = replace(CD_inc, CD_inc == 0, 0),
           UC_inc = replace(UC_inc, UC_inc == 0, 0),
           CD_prev = replace(CD_prev, CD_prev == 0, 0),
           UC_prev = replace(UC_prev, UC_prev == 0, 0),
           CD_change_inc = replace(CD_change_inc, CD_change_inc == 0, 0),
           UC_change_inc = replace(UC_change_inc, UC_change_inc == 0, 0),
           CD_change_prev = replace(CD_change_prev, CD_change_prev == 0, 0),
           UC_change_prev = replace(UC_change_prev, UC_change_prev == 0, 0),
           CD_inc_impute = NA,
           UC_inc_impute = NA, 
           CD_prev_impute = NA,
           UC_prev_impute = NA,
           CD_change_inc_impute = NA,
           UC_change_inc_impute = NA, 
           CD_change_prev_impute = NA,
           UC_change_prev_impute = NA) %>%
    select(-stage_inc, -stage_prev, -decade.x, -decade.y, -status_inc, -status_prev)

# Split dataset into country-decade subsets for data imputation
data_sp <- split(combined, combined$country) 

# Create empty data frames for all disease x data type subsets for imputations
# List of prefixes and suffixes for the data frames
prefixes <- c("CDinc", "UCinc", "CDprev", "UCprev", "CDchange_inc", "UCchange_inc", "CDchange_prev", "UCchange_prev")
suffixes <- 1:3 

# Loop over each prefix + suffix combination
for (prefix in prefixes) {
  for (suffix in suffixes) {
    data_frame_name <- paste0(prefix, suffix)
    assign(data_frame_name, data.frame())
  }
}

# Impute missing data so that there are no NA values for any disease x data-type subset
    # incidence, prevalence, and rates of change are imputed
# For loop to iterate over each country-decade split to impute missing values
    # output of for loop is complete data frame
# If possible, linear interpolation is used
    # --> if linear interpolation conditions not met, mean imputation is used
        # --> if no data available for mean calculation, zero imputation is used

for (i in 1:length(data_sp)) {
        
    if (nrow(data_sp[[i]]) > 0) {
        
        # Create jittered year and order by year_data
        set.seed(5678)
        data_sp[[i]] <- data_sp[[i]] %>% 
            complete(year_data = seq(min(year_data), max(year_data), by = 1)) %>%
            mutate(year_jitter = jitter(year_data, factor = 1e-1)) %>%
            arrange(year_data) # use year_data instead of year_jitter for ordering
        
        # CD inc impute
        
        # zero impute if all available data are NA
        if (length(which(is.na(data_sp[[i]]$CD_inc))) == nrow(data_sp[[i]])){
            
            data_sp[[i]]$CD_inc_impute <- ifelse(is.na(data_sp[[i]]$CD_inc), 0, data_sp[[i]]$CD_inc)
            CDinc1 <- rbind(CDinc1, data_sp[[i]])
            
        } else {
            
            # linear interpolate if there are more than 2 non-NA values
            if (length(which(!is.na(data_sp[[i]]$CD_inc))) >= 2) {
                
                CDinc_lin_fun <- approxfun(x = data_sp[[i]]$year_jitter,
                                           y = data_sp[[i]]$CD_inc,
                                           method = "linear",
                                           rule = 2) # extrapolate to remove NA
                data_sp[[i]]$CD_inc_impute <- CDinc_lin_fun(data_sp[[i]]$year_jitter)
                CDinc2 <- rbind(CDinc2, data_sp[[i]])
                
            # mean imputation elsewhere        
            } else {
                
                data_sp[[i]]$CD_inc_impute <- ifelse(is.na(data_sp[[i]]$CD_inc), mean(data_sp[[i]]$CD_inc, na.rm = TRUE), data_sp[[i]]$CD_inc)
                CDinc3 <- rbind(CDinc3, data_sp[[i]])
                
            }
        }
        
        CDincimpute <- rbind(CDinc1, CDinc2, CDinc3)
        
        # UC inc impute
        
        # zero impute if all available data are NA
        if (length(which(is.na(data_sp[[i]]$UC_inc))) == nrow(data_sp[[i]])){
            
            data_sp[[i]]$UC_inc_impute <- ifelse(is.na(data_sp[[i]]$UC_inc), 0, data_sp[[i]]$UC_inc)
            UCinc1 <- rbind(UCinc1, data_sp[[i]])
            
        } else {
            
            # linear interpolate if there are more than 2 non-NA values
            if (length(which(!is.na(data_sp[[i]]$UC_inc))) >= 2) {
                
                UCinc_lin_fun <- approxfun(x = data_sp[[i]]$year_jitter,
                                           y = data_sp[[i]]$UC_inc,
                                           method = "linear",
                                           rule = 2) # extrapolate to remove NA
                data_sp[[i]]$UC_inc_impute <- UCinc_lin_fun(data_sp[[i]]$year_jitter)
                UCinc2 <- rbind(UCinc2, data_sp[[i]])
            
            # mean imputation elsewhere        
            } else {
                
                data_sp[[i]]$UC_inc_impute <- ifelse(is.na(data_sp[[i]]$UC_inc), mean(data_sp[[i]]$UC_inc, na.rm = TRUE), data_sp[[i]]$UC_inc)
                UCinc3 <- rbind(UCinc3, data_sp[[i]])
                
            }
        }
        
        UCincimpute <- rbind(UCinc1, UCinc2, UCinc3)
        
        # CD prev impute
        
        # zero impute if all available data are NA
        if (length(which(is.na(data_sp[[i]]$CD_prev))) == nrow(data_sp[[i]])){
            
            data_sp[[i]]$CD_prev_impute <- ifelse(is.na(data_sp[[i]]$CD_prev), 0, data_sp[[i]]$CD_prev)
            CDprev1 <- rbind(CDprev1, data_sp[[i]])
            
        } else {
            
           # linear interpolate if there are more than 2 non-NA values
            if (length(which(!is.na(data_sp[[i]]$CD_prev))) >= 2) {
                
                CDprev_lin_fun <- approxfun(x = data_sp[[i]]$year_jitter,
                                            y = data_sp[[i]]$CD_prev,
                                            method = "linear",
                                            rule = 2) # extrapolate to remove NA
                data_sp[[i]]$CD_prev_impute <- CDprev_lin_fun(data_sp[[i]]$year_jitter)
                CDprev2 <- rbind(CDprev2, data_sp[[i]])
            
            # mean imputation elsewhere        
            } else {
                
                data_sp[[i]]$CD_prev_impute <- ifelse(is.na(data_sp[[i]]$CD_prev), mean(data_sp[[i]]$CD_prev, na.rm = TRUE), data_sp[[i]]$CD_prev)
                CDprev3 <- rbind(CDprev3, data_sp[[i]])
                
            }
        }
        
        CDprevimpute <- rbind(CDprev1, CDprev2, CDprev3)
        
        # UC prev impute
        
        # zero impute if all available data are NA
        if (length(which(is.na(data_sp[[i]]$UC_prev))) == nrow(data_sp[[i]])){
            
            data_sp[[i]]$UC_prev_impute <- ifelse(is.na(data_sp[[i]]$UC_prev), 0, data_sp[[i]]$UC_prev)
            UCprev1 <- rbind(UCprev1, data_sp[[i]])
            
        } else {
            
            # linear interpolate if there are more than 2 non-NA values and year gaps of more than 10 years
            if (length(which(!is.na(data_sp[[i]]$UC_prev))) >= 2) {
                
                UCprev_lin_fun <- approxfun(x = data_sp[[i]]$year_jitter,
                                            y = data_sp[[i]]$UC_prev,
                                            method = "linear",
                                            rule = 2) # extrapolate to remove NA
                data_sp[[i]]$UC_prev_impute <- UCprev_lin_fun(data_sp[[i]]$year_jitter)
                UCprev2 <- rbind(UCprev2, data_sp[[i]])
            
            # mean imputation elsewhere        
            } else {
                
                data_sp[[i]]$UC_prev_impute <- ifelse(is.na(data_sp[[i]]$UC_prev), mean(data_sp[[i]]$UC_prev, na.rm = TRUE), data_sp[[i]]$UC_prev)
                UCprev3 <- rbind(UCprev3, data_sp[[i]])
                
            }
        }
        
        UCprevimpute <- rbind(UCprev1, UCprev2, UCprev3)
        
        # CD change inc impute
        
        # zero impute if all available data are NA
        if (length(which(is.na(data_sp[[i]]$CD_change_inc))) == nrow(data_sp[[i]])){
            
            data_sp[[i]]$CD_change_inc_impute <- ifelse(is.na(data_sp[[i]]$CD_change_inc), 0, data_sp[[i]]$CD_change_inc)
            CDchange_inc1 <- rbind(CDchange_inc1, data_sp[[i]])
            
        } else {
            
            # linear interpolate if there are more than 2 non-NA values
            if (length(which(!is.na(data_sp[[i]]$CD_change_inc))) >= 2) {
                
                CDchange_inc_lin_fun <- approxfun(x = data_sp[[i]]$year_jitter,
                                                  y = data_sp[[i]]$CD_change_inc,
                                                  method = "linear",
                                                  rule = 2) # extrapolate to remove NA
                data_sp[[i]]$CD_change_inc_impute <- CDchange_inc_lin_fun(data_sp[[i]]$year_jitter)
                CDchange_inc2 <- rbind(CDchange_inc2, data_sp[[i]])
            
            # mean imputation elsewhere        
            } else {
                
                data_sp[[i]]$CD_change_inc_impute <- ifelse(is.na(data_sp[[i]]$CD_change_inc), mean(data_sp[[i]]$CD_change_inc, na.rm = TRUE), data_sp[[i]]$CD_change_inc)
                CDchange_inc3 <- rbind(CDchange_inc3, data_sp[[i]])
                
            }
        }
        
        CDchange_inc_impute <- rbind(CDchange_inc1, CDchange_inc2, CDchange_inc3)
        
        # UC change inc impute
        
        # zero impute if all available data are NA
        if (length(which(is.na(data_sp[[i]]$UC_change_inc))) == nrow(data_sp[[i]])){
            
            data_sp[[i]]$UC_change_inc_impute <- ifelse(is.na(data_sp[[i]]$UC_change_inc), 0, data_sp[[i]]$UC_change_inc)
            UCchange_inc1 <- rbind(UCchange_inc1, data_sp[[i]])
            
        } else {
            
            # linear interpolate if there are more than 2 non-NA values
            if (length(which(!is.na(data_sp[[i]]$UC_change_inc))) >= 2) {
                
                UCchange_inc_lin_fun <- approxfun(x = data_sp[[i]]$year_jitter,
                                                  y = data_sp[[i]]$UC_change_inc,
                                                  method = "linear",
                                                  rule = 2) # extrapolate to remove NA
                data_sp[[i]]$UC_change_inc_impute <- UCchange_inc_lin_fun(data_sp[[i]]$year_jitter)
                UCchange_inc2 <- rbind(UCchange_inc2, data_sp[[i]])
            
            # mean imputation elsewhere        
            } else {
                
                data_sp[[i]]$UC_change_inc_impute <- ifelse(is.na(data_sp[[i]]$UC_change_inc), mean(data_sp[[i]]$UC_change_inc, na.rm = TRUE), data_sp[[i]]$UC_change_inc)
                UCchange_inc3 <- rbind(UCchange_inc3, data_sp[[i]])
                
            }
        }
        
        UCchange_inc_impute <- rbind(UCchange_inc1, UCchange_inc2, UCchange_inc3)
        
        # CD change prev impute
        
        # zero impute if all available data are NA
        if (length(which(is.na(data_sp[[i]]$CD_change_prev))) == nrow(data_sp[[i]])){
            
            data_sp[[i]]$CD_change_prev_impute <- ifelse(is.na(data_sp[[i]]$CD_change_prev), 0, data_sp[[i]]$CD_change_prev)
            CDchange_prev1 <- rbind(CDchange_prev1, data_sp[[i]])
            
        } else {
            
            # linear interpolate if there are more than 2 non-NA values
            if (length(which(!is.na(data_sp[[i]]$CD_change_prev))) >= 2) {
                
                CDchange_prev_lin_fun <- approxfun(x = data_sp[[i]]$year_jitter,
                                                   y = data_sp[[i]]$CD_change_prev,
                                                   method = "linear",
                                                   rule = 2) # extrapolate to remove NA
                data_sp[[i]]$CD_change_prev_impute <- CDchange_prev_lin_fun(data_sp[[i]]$year_jitter)
                CDchange_prev2 <- rbind(CDchange_prev2, data_sp[[i]])
            
            # mean imputation elsewhere        
            } else {
                
                data_sp[[i]]$CD_change_prev_impute <- ifelse(is.na(data_sp[[i]]$CD_change_prev), mean(data_sp[[i]]$CD_change_prev, na.rm = TRUE), data_sp[[i]]$CD_change_prev)
                CDchange_prev3 <- rbind(CDchange_prev3, data_sp[[i]])
                
            }
        }
        
        CDchange_prev_impute <- rbind(CDchange_prev1, CDchange_prev2, CDchange_prev3)
        
        # UC change prev impute
        
        # zero impute if all available data are NA
        if (length(which(is.na(data_sp[[i]]$UC_change_prev))) == nrow(data_sp[[i]])){
            
            data_sp[[i]]$UC_change_prev_impute <- ifelse(is.na(data_sp[[i]]$UC_change_prev), 0, data_sp[[i]]$UC_change_prev)
            UCchange_prev1 <- rbind(UCchange_prev1, data_sp[[i]])
            
        } else {
            
            # linear interpolate if there are more than 2 non-NA values
            if (length(which(!is.na(data_sp[[i]]$UC_change_prev))) >= 2) {
                
                UCchange_prev_lin_fun <- approxfun(x = data_sp[[i]]$year_jitter,
                                                   y = data_sp[[i]]$UC_change_prev,
                                                   method = "linear",
                                                   rule = 2) # extrapolate to remove NA
                data_sp[[i]]$UC_change_prev_impute <- UCchange_prev_lin_fun(data_sp[[i]]$year_jitter)
                UCchange_prev2 <- rbind(UCchange_prev2, data_sp[[i]])
            
            # mean imputation elsewhere        
            } else {
                
                data_sp[[i]]$UC_change_prev_impute <- ifelse(is.na(data_sp[[i]]$UC_change_prev), mean(data_sp[[i]]$UC_change_prev, na.rm = TRUE), data_sp[[i]]$UC_change_prev)
                UCchange_prev3 <- rbind(UCchange_prev3, data_sp[[i]])
                
            }
        }
        
        UCchange_prev_impute <- rbind(UCchange_prev1, UCchange_prev2, UCchange_prev3)
       
    } else {
    
    }
    
} 

# Create imputed data frame

imputed_full <- as.data.frame(UCchange_prev_impute)
row.names(imputed_full) <- paste0(rownames(imputed_full), "_", imputed_full$country, "_", imputed_full$name, "_", imputed_full$year_data) 

# Zero-impute any remaining NAs in feature columns
# Zero-impute any negative values for incidence and prevalence resulting from linear interpolation
# Add imputed indicator columns

imputed_full <- imputed_full %>%
    mutate_at(c("CD_inc_impute", "UC_inc_impute", "CD_prev_impute", "UC_prev_impute", 
                "CD_change_inc_impute", "UC_change_inc_impute", "CD_change_prev_impute", "UC_change_prev_impute"), 
              ~replace_na(., 0)) %>%
    mutate(across(c("CD_inc_impute", "UC_inc_impute", "CD_prev_impute", "UC_prev_impute"), # remove all negative imputed values from inc and prev
                  ~ ifelse(. < 0, 0, .))) %>% 
    mutate(CD_inc_ind = is.na(CD_inc),
           UC_inc_ind = is.na(UC_inc),
           CD_prev_ind = is.na(CD_prev),
           UC_prev_ind = is.na(UC_prev),
           CD_change_inc_ind = is.na(CD_change_inc),
           UC_change_inc_ind = is.na(UC_change_inc),
           CD_change_prev_ind = is.na(CD_change_prev),
           UC_change_prev_ind = is.na(UC_change_prev)) %>%
    filter(!is.na(country)) %>%
    select(-year_jitter)

```

#### 5.1 Imputation Viz
```{r, warning = FALSE}
# split dataframe by country
impute_sp <- split(imputed_full, imputed_full$country)

for(i in 1:length(impute_sp)) {
    
    p1 <- ggplot(impute_sp[[i]]) +
        geom_point(aes(year_data, CD_inc_impute), colour = "darkred") +
        geom_point(aes(year_data, CD_inc), colour = "black")+ ggtitle(paste0("CD Incidence ", unique(impute_sp[[i]]$country)))+
        ylim(0,NA) + theme_bw()
    
    p2 <- ggplot(impute_sp[[i]]) +
        geom_point(aes(year_data, UC_inc_impute), colour = "darkred") +
        geom_point(aes(year_data, UC_inc), colour = "black")+ ggtitle("UC Incidence")+
        ylim(0,NA) + theme_bw()
    
    p3 <- ggplot(impute_sp[[i]]) +
        geom_point(aes(year_data, CD_prev_impute), colour = "darkred") +
        geom_point(aes(year_data, CD_prev), colour = "black")+ ggtitle("CD Prevalence")+
        ylim(0,NA) + theme_bw()
    
    p4 <- ggplot(impute_sp[[i]]) +
        geom_point(aes(year_data, UC_prev_impute), colour = "darkred") +
        geom_point(aes(year_data, UC_prev), colour = "black")+ ggtitle("UC Prevalence")+
        ylim(0,NA) + theme_bw()
    
    p <- ggarrange(p1, p2, p3, p4,
             ncol=2, nrow = 2, common.legend = TRUE, legend="bottom")  
    #annotate_figure(p, top = text_grob(data_sp[[i]]$country), face = "bold", size = 14))
    print(p)
    
}
```

## 6. Random forest

### 6a. Data split
```{r rf data split}

# Create training/validation data set

train_validate <- imputed_full %>%
    filter(status == "labelled")

train_validate_labels <- train_validate %>% 
    select(stage)

train_validate_context <- train_validate %>%
    select(-stage)

# Create test (i.e., unclassified) data set
# Stage classifications to be assigned with model fit using training and validation data

test <- imputed_full %>%
    filter(status == "unlabelled")
test$stage <- as.factor(test$stage)

test_labels <- test %>% 
    select(stage)

test_context <- test %>%
    select(-stage)

```

### 6b. Training and validating
```{r rf train and validate}

# Create training and validation sets and train model

set.seed(1234)

sample_index <- sample(nrow(train_validate), round(nrow(train_validate)*.75), replace = FALSE)

train <- train_validate[sample_index, ]
train$stage <- as.factor(train$stage)

validation <- train_validate[-sample_index, ] # all the rows not given a sample index
validation$stage <- as.factor(validation$stage)

train_labels <- train_validate_labels[sample_index, ] # may need to wrap in as.factor for non-dataframe
validation_labels <- train_validate_labels[-sample_index, ]

train_context <- train_validate_context[sample_index, ] # may need to wrap in as.factor for non-dataframe
validation_context <- train_validate_context[-sample_index, ]

# Hyperparameter tuning

# Define the parameter grid
mtry_values <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16)
ntree_values <- c(500, 1000, 1500, 2000)

# Create a data frame to store results
tune_results <- data.frame(mtry = integer(), ntree = integer(), Accuracy = numeric())

set.seed(1234)

# Loop through combinations
for (mtry in mtry_values) {
  for (ntree in ntree_values) {
    rf_tune <- randomForest(formula = stage ~ CD_inc_impute + CD_prev_impute + UC_inc_impute + UC_prev_impute + CD_change_inc_impute + CD_change_prev_impute + UC_change_inc_impute + UC_change_prev_impute + CD_inc_ind + UC_inc_ind + CD_prev_ind + UC_prev_ind + CD_change_inc_ind + UC_change_inc_ind + CD_change_prev_ind + UC_change_prev_ind,
                       data = train,
                       ntree = ntree, # for imputed across geo and country all-time ntree = 1000
                       na.action = na.omit,
                       mtry = mtry, # for imputed across geo all-time mtry = 6, country all-time = 4
                       importance = TRUE)
    oob <- rf_tune$err.rate[ntree, "OOB"]
    confusion <- rf_tune$confusion
    class_error <- confusion[, "class.error"]
    pred <- predict(rf_tune, validation[,15:30])
    acc <- mean(pred == validation$stage)
    tune_results <- rbind(tune_results, data.frame(mtry = mtry, ntree = ntree, accuracy = acc, OOB = oob, err_stage1 = class_error[1], err_stage2 = class_error[2], err_stage3 = class_error[3]))
  }
}

# Display results
print(tune_results)

OOB_check <- tune_results %>%
    slice(which.min(OOB))

acc_check <- tune_results %>%
    slice(which.max(accuracy))

# Create object rf with final random forest model
# mtry and ntree values determined by above hyperparameter tuning
# final model ntree = 1000, mtry = 5
set.seed(1234)
rf <- randomForest(formula = stage ~ CD_inc_impute + CD_prev_impute + UC_inc_impute + UC_prev_impute + CD_change_inc_impute + CD_change_prev_impute + UC_change_inc_impute + UC_change_prev_impute + CD_inc_ind + UC_inc_ind + CD_prev_ind + UC_prev_ind + CD_change_inc_ind + UC_change_inc_ind + CD_change_prev_ind + UC_change_prev_ind, 
                   data = train, 
                   ntree = 1000, 
                   na.action = na.omit, 
                   mtry = 5, 
                   importance = TRUE)
rf

# Extract feature importance measures from model
rf_imp <- as.data.frame(rf$importance)
rf_imp$measure <- row.names(rf_imp)
imp_inc_prev <- rf_imp[1:4,]

rf_imp_plot <- plot_ly(data = imp_inc_prev, 
                           x = ~measure) %>%
    add_trace(y = ~stage1,
              name = "Stage 1",
              type = "bar",
              marker = list(color = "#7570B3")) %>%
    add_trace(y = ~stage2,
              name = "Stage 2",
              type = "bar",
              marker = list(color = "#D95F02")) %>%
    add_trace(y = ~stage3,
              name = "Stage 3",
              type = "bar",
              marker = list(color = "#1B9E77")) %>%
    layout(xaxis = list(title = list(text = "Incidence or prevalence feature type",
                                     standoff = 30),
                        categoryorder = "array",
                        categoryarray = c("CD_inc_impute", "UC_inc_impute", "CD_prev_impute", "UC_prev_impute"),
                        tickvals = c("CD_inc_impute", "UC_inc_impute", "CD_prev_impute", "UC_prev_impute"),
                        ticktext = c("CD\nincidence", "UC\nincidence", "CD\nprevalence", "UC\nprevalence"), 
                        tickangle = 0,
                        titlefont = list(size = 35),
                        tickfont = list(size = 30)),
           yaxis = list(title = list(text = "Mean decrease in accuracy (%)",
                                     standoff = 30),
                        tickvals = c(0, 0.05, 0.1, 0.15, 0.2, 0.25, 0.3, 0.35, 0.4, 0.45, 0.5),
                        ticktext = c("0", "0.05", "0.10", "0.15", "0.20", "0.25", "0.30", "0.35", "0.40", "0.45", "0.50"),
                        titlefont = list(size = 35),
                        tickfont = list(size = 30)), 
           barmode = 'group',
           legend = list(x = 1.01, y = 0.5,
                         font = list(size = 35)))

# rf_imp_plot %>% htmlwidgets::onRender(
#   "function(el, x) {
#   var gd = document.getElementById(el.id);
#   Plotly.downloadImage(gd, {format: 'png', width: 1750, height: 2000, filename: 'rf_importance_2024-10-10', scale: 5});
#   }"
#  )

# Random forest model stage prediction for validation data
validation_prediction <- predict(rf, validation[,15:30])
validation_rf_results <- confusionMatrix(data = validation_prediction, reference = as.factor(validation$stage)) # model accuracy here
validation_tabulation <- table(validation_prediction, as.factor(validation$stage))

```

### 6c. Classify unlabelled data

```{r rf test}

# Random forest model stage prediction for test (i.e., unlabelled) data
set.seed(1234)
test_prediction <- data.frame(predict(rf, test[,15:30]))

# Create dataframe with test data and corresponding stage predictions

test_complete <- data.frame(test, test_prediction)

names(test_complete)[31] <- "predicted_stage" # rename predicted stage column

train_validate <- train_validate %>%
    mutate("predicted_stage" = stage)

```

### 6d. Output
```{r rf output}

# Function to calculate the mode
get_mode <- function(x) {
  unique_x <- unique(x)
  unique_x[which.max(tabulate(match(x, unique_x)))]
}

# Combine training,  validation, and test data

output <- rbind(train_validate, test_complete)

output_annual <- output %>%
    arrange(country) %>%
    select(country, name, year_data, decade, CD_inc, UC_inc, CD_prev, UC_prev, predicted_stage, status)

output_decade <- output %>%
    group_by(country, decade) %>%
    summarize(predicted_stage = get_mode(predicted_stage))

output_org <- output %>%
    group_by(country, decade) %>%
    summarize(collapsed = get_mode(predicted_stage)) %>%
    mutate(stage_val = case_when(collapsed == "stage1" ~ 1, collapsed == "stage2" ~ 2, collapsed == "stage3" ~ 3),
           sequence = seq_along(row_number())) %>%
    mutate(stage_val_final = ifelse(sequence > 1 & stage_val < cummax(stage_val), cummax(stage_val), stage_val),
           greater = stage_val < cummax(stage_val)) %>%
    mutate(predicted_stage = case_when(stage_val_final == 1 ~ "stage1", stage_val_final == 2 ~ "stage2", stage_val_final == 3 ~ "stage3")) %>%
    select(country, decade, predicted_stage)

# Write random forest output to CSV
# write.csv(output_annual, "~/file/path/your_output.csv", row.names = FALSE) # leave line commented out when finished with script

```