---
title: "4-stages-IBD Regional Scatterplots"
author: "Julia Gorospe"
date: "20232-12-12"
output: html_document
R version: 4.2.1 (2022-06-23)
Platform: x86_64-apple-darwin17.0 (64-bit)
Running under: macOS Monterey 12.5.1
---

# 1 Script Setup
1.1 Resetting environment
```{r, setup, include = FALSE}
# Clear environment
rm(list = ls(all = TRUE))

# Set directory
require("knitr")
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/Dropbox/Ratio_2021 (2)')   #CHANGE ME!

# Activate reticulate
library(reticulate)
virtualenv_create("r-reticulate")
#virtualenv_install("kaleido", envname = "r-reticulate")
#virtualenv_install("python-poppler", envname = "r-reticulate")
use_virtualenv("r-reticulate")

# Install and load packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rio,                 # v0.5.16 import/exporting files
               tidyverse,           # v1.3.0
               plotly,
               htmlwidgets,
               fANCOVA)
```

1.2 Load data
```{r, warning = FALSE}
raw <- import("https://raw.githubusercontent.com/kaplan-gi/4-stages-IBD/main/1_Data/GIVES21_example_data.csv")
```

# 2 Data Prep
2.1 Filter data
```{r}
# Tidy dataset and back calculate missing populations
data <- raw %>%
  filter(age_cat == "All")
```

2.2 Subset data
```{r}
# Incidence datasets
country_listI <- data %>%
  filter(data_type == "Incidence") %>%
  split(., f = .$country)                  # separate df for each country

# Prevalence datasets
country_listP <- data %>%
  filter(data_type == "Prevalence") %>%
  split(., f = .$country)         
```

# 3 Plot Design
3.1 Formatting
```{r}
# margin
marg <- list(l = 75, r = 75,
               t = 25, b = 25)

# colors
green <- c("#1B9E77")
grey <- c("#666666")
orange <- c("#D95F02")
purple <- c("#7570B3")


# x-axis ranges
xrange <- data %>% 
  group_by(country) %>% 
  summarise(xmin = min(year_data) - 1,
            xmax = max(year_data) + 1)

# stacked incidence scatterplot
A <- list(x = -0.09,
          y = 1.1,
          text = "<b>A</b>",
          xref = "paper",
          yref = "paper",
          xanchor = "center",
          yanchor = "middle",
          showarrow = FALSE,
          font = list(size = 18))

B <- list(x = -0.09,
          y = 1.1,
          text = "<b>B</b>",
          xref = "paper",
          yref = "paper",
          xanchor = "center",
          yanchor = "middle",
          showarrow = FALSE,
          font = list(size = 18))
```

# 4 Plot Generation
4.1 Generating incidence boxplots for each country
```{r, warning = FALSE}
boxesI <- lapply(country_listI, function (x)
  plot <- x %>% 
    plot_ly(type = "box",
            boxpoints = "all",
            jitter = 0.8,
            pointpos = 0,
            showlegend = FALSE) %>% 
    add_trace(y = ~CD_rate,
              color = I(grey),
              text = paste0("CD Rate: ", round(x$CD_rate,digits = 2), 
                            "<br>Year: ", x$year_data,
                            "<br>Study: ", x$author, " ", x$year_study, 
                            "<br>Location: ", x$name),
              hoverinfo = "text",
              name = "CD") %>% 
    add_trace(y = ~UC_IBDu_rate,
              color = I(green),
              text = paste0("UC + IBD-u Rate: ", round(x$UC_IBDu_rate,digits = 2),
                            "<br>Year: ", x$year_data,
                            "<br>Study: ", x$author, " ", x$year_study, 
                            "<br>Location: ", x$name),
              hoverinfo = "text",
              name = "UC + IBD-u") %>% 

    layout(xaxis = list(title = "Disease"),
           yaxis = list(range = list(0, (1.1*(max(x$UC_IBDu_rate, x$CD_rate, na.rm = TRUE))))),
           margin = marg)
  )
```

4.2 Generating prevalence boxplots for each country
```{r, warning = FALSE}
boxesP <- lapply(country_listP, function (x)
  p <- x %>% 
    plot_ly(type = "box",
            boxpoints = "all",
            jitter = 0.8,
            pointpos = 0,
            showlegend = FALSE) %>% 
    add_trace(y = ~CD_rate,
              color = I(purple),
              text = paste0("CD Rate: ", round(x$CD_rate,digits = 2),
                            "<br>Year: ", x$year_data,
                            "<br>Study: ", x$author, " ", x$year_study,
                            "<br>Location: ", x$name),
              hoverinfo = "text",
              name = "CD") %>% 
    add_trace(y = ~UC_IBDu_rate,
              color = I(orange),
              text = paste0("UC + IBD-u Rate: ", round(x$UC_IBDu_rate, digits = 2),
                            "<br>Year: ", x$year_data,
                            "<br>Study: ", x$author, " ", x$year_study, 
                            "<br>Location: ", x$name),
            hoverinfo = "text",
            name = "UC + IBD-u") %>% 
    layout(xaxis = list(title = "Disease"),
           yaxis = list(range = list(0, (1.1*(max(x$UC_IBDu_rate, x$CD_rate, na.rm = TRUE))))),
         margin = marg))
```

4.3 Generating incidence scatterplots for each country
```{r, warning = FALSE}
scatterI <- lapply(country_listI, function (x) {
  
    # xrange
    xmin <- xrange$xmin[xrange$country == unique(x$country)] 
    xmax <- xrange$xmax[xrange$country == unique(x$country)]
  
    # plot
    p <- x %>% 
      plot_ly(type = "scatter") %>% 
      add_markers(x = ~year_data, y = ~CD_rate,
                  mode = "markers",
                  opacity = 0.8,
                  marker = list(color = grey),
                  text = paste0("CD Rate: ", round(x$CD_rate, digits = 2),
                                "<br>Year: ", x$year_data,
                                "<br>Study: ", x$author, " ", x$year_study,
                                "<br>Location: ", x$name),
                  hoverinfo = "text",
                  name = "CD Incidence",
                  legendgroup = "A") %>% 
      add_markers(x = ~year_data, y = ~UC_IBDu_rate,
                  mode = "markers",
                  opacity = 0.8,
                  marker = list(color = green),
                  text = paste0("UC + IBD-u Rate: ", round(x$UC_IBDu_rate, digits = 2),
                                "<br>Year: ", x$year_data,
                                "<br>Study: ", x$author, " ", x$year_study,
                                "<br>Location: ", x$name),
                  hoverinfo = "text",
                  name = "UC + IBD-u Incidence",
                  legendgroup = "A") %>% 
      layout(xaxis = list(title = "Year",
                          range = list(xmin, xmax)),
             yaxis = list(title = "Incidence (per 100,000)",
                          range = list(0, (1.1*(max(x$UC_IBDu_rate, x$CD_rate, na.rm = TRUE))))),
             annotations = A,
             margin = marg)
    
    # add loess trend lines if there are enough datapoints
    if (length(unique(x$UC_IBDu_rate)) >= 5 & length(unique(x$year_data)) >= 5) {
      noNA_UC <- x %>% filter(!is.na(UC_IBDu_rate))
      # if there is a density of data of greater than 1 per year then use auto if not use default 0.75
      if (nrow(noNA_UC)/length(unique(noNA_UC$year_data)) > 1.25) {
        loess_UC <- loess.as(noNA_UC$year_data, noNA_UC$UC_IBDu_rate, degree = 2, criterion = "aicc",
                           family = "gaussian", user.span = NULL)
      } else {
        loess_UC <- loess.as(noNA_UC$year_data, noNA_UC$UC_IBDu_rate, degree = 2, criterion = "aicc",
                           family = "gaussian", user.span = 0.8)
      }

      loess_curve_UC <- curve(predict(loess_UC, newdata = data.frame(x = x)), 
                              from  = min(noNA_UC$year_data), to = max(noNA_UC$year_data))
      p <- p %>% 
        add_trace(x = ~loess_curve_UC$x, y = ~loess_curve_UC$y,
                  mode = "lines",
                  opacity = 0.4,
                  line = list(color = green, width= 4),
                  hoverinfo = "skip",
                  showlegend = FALSE)
    } else {
      p <- p
    }
    
    if (length(unique(x$CD_rate)) >= 5 & length(unique(x$year_data)) >= 5) {
       noNA_CD <- x %>% filter(!is.na(CD_rate))
       
       if (nrow(noNA_CD)/length(unique(noNA_CD$year_data)) > 1.25) {
         loess_CD <- loess.as(noNA_CD$year_data, noNA_CD$CD_rate, degree = 2, criterion = "aicc",
                           family = "gaussian", user.span = NULL)
       } else {
         loess_CD <- loess.as(noNA_CD$year_data, noNA_CD$CD_rate, degree = 2, criterion = "aicc",
                            family = "gaussian", user.span = 0.8)
       }
       
       loess_curve_CD <- curve(predict(loess_CD, newdata = data.frame(x = x)),
                               from  = min(noNA_CD$year_data), to = max(noNA_CD$year_data))
       p <- p %>%
         add_trace(x = ~loess_curve_CD$x, y = ~loess_curve_CD$y,
                   mode = "lines",
                   opacity = 0.4,
                   line = list(color = grey, width = 4),
                   hoverinfo = "skip",
                   showlegend = FALSE)
     } else {
       p <- p
     }
  }
)
```

4.5 Generating prevalence scatterplots for each country
```{r}
scatterP <- lapply(country_listP, function (x) {
  
  # xrange
  xmin <- xrange$xmin[xrange$country == unique(x$country)] 
  xmax <- xrange$xmax[xrange$country == unique(x$country)]
  
  # plot
  p <- x %>% 
    plot_ly(type = "scatter") %>% 
    add_markers(x = ~year_data, y = ~CD_rate,
                mode = "markers",
                opacity = 0.8,
                marker = list(color = purple),
                text = paste0("CD Rate: ", round(x$CD_rate, digits = 2),
                              "<br>Year: ", x$year_data,
                              "<br>Study: ", x$author, " ", x$year_study,
                              "<br>Location: ", x$name),
                hoverinfo = "text",
                name = "CD Prevalence",
                legendgroup = "B") %>% 
    add_markers(x = ~year_data, y = ~UC_IBDu_rate,
                mode = "markers",
                opacity = 0.8,
                marker = list(color = orange),
                text = paste0("UC + IBD-u Rate: ", round(x$UC_IBDu_rate, digits = 2),
                              "<br>Year: ", x$year_data,
                              "<br>Study: ", x$author, " ", x$year_study,
                              "<br>Location: ", x$name),
                hoverinfo = "text",
                name = "UC + IBD-u Prevalence",
                legendgroup = "B") %>% 
    layout(xaxis = list(title = "Year",
                        range = list(xmin, xmax)),
           yaxis = list(title = "Prevalence (per 100,000)",
                        range = list(0, (1.1*(max(x$UC_IBDu_rate, x$CD_rate, na.rm = TRUE))))),
           annotations = B,
           margin = marg)
  
 # add loess trend lines if there are enough datapoints
    if (length(unique(x$UC_IBDu_rate)) >= 5 & length(unique(x$year_data)) >= 5) {
      noNA_UC <- x %>% filter(!is.na(UC_IBDu_rate))
      # if there is a density of data of greater than 1 per year then use auto if not use default 0.75
      if (nrow(noNA_UC)/length(unique(noNA_UC$year_data)) > 1.25) {
        loess_UC <- loess.as(noNA_UC$year_data, noNA_UC$UC_IBDu_rate, degree = 2, criterion = "aicc",
                           family = "gaussian", user.span = NULL)
      } else {
        loess_UC <- loess.as(noNA_UC$year_data, noNA_UC$UC_IBDu_rate, degree = 2, criterion = "aicc",
                           family = "gaussian", user.span = 0.8)
      }

      loess_curve_UC <- curve(predict(loess_UC, newdata = data.frame(x = x)), 
                              from  = min(noNA_UC$year_data), to = max(noNA_UC$year_data))
      p <- p %>% 
        add_trace(x = ~loess_curve_UC$x, y = ~loess_curve_UC$y,
                  mode = "lines",
                  opacity = 0.4,
                  line = list(color = orange, width= 4),
                  hoverinfo = "skip",
                  showlegend = FALSE)
    } else {
      p <- p
      
    }
    
    if (length(unique(x$CD_rate)) >= 5 & length(unique(x$year_data)) >= 5) {
       noNA_CD <- x %>% filter(!is.na(CD_rate))
       
       if (nrow(noNA_CD)/length(unique(noNA_CD$year_data)) > 1.25) {
         loess_CD <- loess.as(noNA_CD$year_data, noNA_CD$CD_rate, degree = 2, criterion = "aicc",
                           family = "gaussian", user.span = NULL)
       } else {
         loess_CD <- loess.as(noNA_CD$year_data, noNA_CD$CD_rate, degree = 2, criterion = "aicc",
                            family = "gaussian", user.span = 0.8)
       }
       
       loess_curve_CD <- curve(predict(loess_CD, newdata = data.frame(x = x)),
                               from  = min(noNA_CD$year_data), to = max(noNA_CD$year_data))
       p <- p %>%
         add_trace(x = ~loess_curve_CD$x, y = ~loess_curve_CD$y,
                   mode = "lines",
                   opacity = 0.4,
                   line = list(color = purple, width = 4),
                   hoverinfo = "skip",
                   showlegend = FALSE)
     } else {
       p <- p
     }
  }
)
```

# 5 Aggregating and Saving
5.1 Aggregating incidence plots
```{r, echo = FALSE}
# Nest lists of plots
plots_listI <- list(scatterI, boxesI)

# Extract first item of each element and assign to new list
plots_listI2 <- list()
for (i in 1:length(plots_listI[[1]])) {
  plots_listI2[[i]] <- lapply(plots_listI, `[[`, i)
}

# Arrange lists in single plot
combinedI <- list()
for (i in 1:length(plots_listI[[1]])) {
  combinedI[[i]] <- subplot(plots_listI2[[i]],
                            widths = c(0.7, 0.3),
                            titleY = TRUE, titleX = TRUE,
                            shareY = TRUE,
                            margin = 0.02) %>% 
    layout(margin = list(b = 50))
}

# Name plots by country
namesI <- names(plots_listI[[1]])
names(combinedI) <- namesI
```

5.2 Aggregating prevalence plots
```{r, warning = FALSE}
# Nest lists of plots
plots_listP <- list(scatterP, boxesP)

# Extract first item of each element and assign to new list
plots_listP2 <- list()
for (i in 1:length(plots_listP[[1]])) {
  plots_listP2[[i]] <- lapply(plots_listP, `[[`, i)
}

# Arrange lists in single plot
combinedP <- list()
for (i in 1:length(plots_listP[[1]])) {
  combinedP[[i]] <- subplot(plots_listP2[[i]],
                            widths = c(0.7, 0.3),
                            titleY = TRUE, titleX = TRUE,
                            shareY = TRUE,
                            margin = 0.02) %>% 
    layout(margin = list(t = 50))
}

# Name plots by country
namesP <- names(plots_listP[[1]])
names(combinedP) <- namesP
```

5.3 Merging incidence and prevalence
```{r}
# List all country names
keys <- unique(c(names(combinedI), names(combinedP)))

# Merge plots from each list by name, if only inc or prev available just keep the single plot
combinedIP <- list()
for (i in keys) {
  if ((i %in% names(combinedI)) & (i %in% names(combinedP))) {
  combinedIP[[i]] <- subplot(combinedI[[i]], combinedP[[i]],
                             nrows = 2,
                             heights = c(0.5,0.5),
                             margin = 0.08,
                             titleY = TRUE, titleX = TRUE) %>% 
    layout(title = paste0("<b>", i, "</b>"),
           legend = list(tracegroupgap = 260,
                         y = 0.5),
           margin = list(t = 75, b = 75))
  } else if ((i %in% names(combinedI)) & !(i %in% names(combinedP))) {
    combinedIP[[i]] <- combinedI[[i]] %>% 
      layout(title = paste0("<b>", i, "</b>"),
             legend = list(y = 0.5),
             margin = list(t = 75, b = 75))
    } else {
      combinedIP[[i]] <- combinedP[[i]] %>% 
        layout(title = paste0("<b>", i, "</b>"),
               legend = list(y = 0.5),
               margin = list(t = 75, b = 75))
      }
}

# Order alphabetically
sortedIP <- combinedIP[order(names(combinedIP))]
```

5.4 Saving
```{r}
#save widgets as self-contained html files, viewed in browser
lapply(names(sortedIP), function(x)
  htmlwidgets::saveWidget(widget = sortedIP[[x]],
                         file = paste0(x, "_Plot.html"),
                         selfcontained = TRUE))


#save plotly objects as static png or pdf images
reticulate::py_run_string("import sys")
lapply(names(sortedIP), function(x)
  save_image(sortedIP[[x]],
             file = paste0(x, "_Plot.png"),
             width = 1000, height = 700,
             scale = 6))

lapply(names(sortedIP), function(x)
  save_image(sortedIP[[x]],
             file = paste0(x, "_Plot.pdf"),
             width = 1000, height = 700,
             scale = 1))   #issue with MathJax 
```
